Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-18 text/ms.tex
Edit a comment
Lots of edits and wrap text

Black swans are statistically improbable events that nonetheless occur---often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects (\InsectaRangePerc\%). Black swans were predominantly downward events (\percBSDown \%) and were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors. Extreme events were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies---particularly as the frequency and magnitude of climate extremes increase over the next century.

@@

Black swans are unexpected extreme events with potentially dramatic consequences \citep{taleb2007,sornette2009}. One of the most striking black swans in ecology is the possible asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with shifts in mean temperature, are expected to cause the greatest ecological and societal damage \citep{ipcc2012}. But, while extremes in the physical environment such as wave height, storm severity, and temperature are frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent to which ecological systems buffer or suffer from black swans \citep{nunez2012}.

There is compelling anecdotal evidence for ecological black swans, but systematic evidence across taxa has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}. In fact, anecdotes of population catastrophes are numerous and catastrophes may be the most important element affecting population persistence \citep{mangel1994}. For marine mammal populations, we have compelling evidence of catastrophes \citep{gerber2001, ward2007}, and recently, heavy-tailed distributions have been discovered in time-series of marine microbe abundance \citep{segura2013} and time to extinction for experimental waterflea populations \citep{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which uncovered little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model represents population growth as a linear function in $\log$ space. If we let $x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
The growth parameter $\lambda$ represents the expected growth rate if $N_t = 1$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependent if $b < 0$. Usually, the process noise $\epsilon_t$ is modelled as normally distributed, but in our paper we assume it is drawn from a t distribution with scale parameter $\sigma$ and degrees of freedom $\nu$. In previous analyses of the GPDD, the Gompertz is most often identified as the most parsimonious population model that can be fit to these data \citep{brook2006}.

@@

We investigated possible covariates of heavy-tailed population dynamics visually and through multilevel modelling. We plotted characteristics of the time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two life-history characteristics (body length and maximum lifespan obtained from \citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu < 10) > 0.5$. We formally investigated these relationships by fitting beta regression multilevel models \citep{ferrari2004}. For covariates that were derived from Gompertz model parameter estimates, we incorporated standard deviations around the means. To account for broad patterns of phylogenetic relatedness, we allowed for hierarchical intercepts at the taxonomic class, order, and species level (Supporting Information). We fit our model in Stan with weakly informative priors on the coefficients \citep{gelman2008d} and variance parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c} \citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).

@@

We found strong, but rare, evidence for black-swan population dynamics. By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).

@@

%TODO IMPORTANT quantify across all identified black swan events how many were downward and how many upward. The assumption of your model is a symmetric distribution, but if in fact downward events are more common, perhaps a skewed distribution would be more realistic. Furthermore, if there is skew this would *increase* the probability of higher extinction risk. Something to definitely highlight in the results and abstract.

@@

Our results clarify previous related analyses. An analysis with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black-swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails. This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black-swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations. Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or an intrinsic change in population variability through time. Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black-swan ecological dynamics. This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans. A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.

Our results suggest a number of research questions related to ecological black swans. Given that black swans do occur, can we forecast the probability of black swans in space and time? Furthermore, what management policies allow us to detect them quickly after they happen? Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment? We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}. Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk---the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.

@@

We thank Justin D. Yeakel, TODO, and other members of the Earth to Ocean research group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Silhouette images were obtained from \texttt{phylopic.org} under Creative Commons licenses; sources are listed in the Supporting Information. Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).

@@

\bibliographystyle{ecologyletters}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer populations in panel a are omitted for space. Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}
\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-t process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter ($1$ is maximally density independent, $0$ is maximally density dependent, and $<0$ is inversely density dependent). Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression multilevel model are shown in panels a and d where there was a high probability the slope coefficient was different from zero (Fig.~\ref{fig:posteriors}b).
}
\label{fig:correlates}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{
Posterior probability distributions from beta regression multilevel models.
(a) Taxonomic-order-level posterior densities of Pr($\nu < 10$) (approximately the probability of heavy tails) after accounting for time-series length. Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps). Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals). Long dotted vertical line in panel a indicates the median expected Pr($\nu < 10$) from the prior distribution.
(b) Main effect posterior densities for potential covariates of Pr($\nu < 10$). The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species. All covariates were standardized by subtracting their mean and dividing by twice their standard deviation.
In both panels, short vertical line segments indicate median posterior estimates.
}
\label{fig:posteriors}
\end{center}
\end{figure}

>>>>

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in
ecological process noise. We apply our model to \NPops~time series from around
the world across \NOrders~taxonomic orders and seven classes. We find strong
evidence of black swans, but they are rare occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

% TODO highlight that skew would *increase* the probability of higher
% extinction risk

@@

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the possible asteroid collision marking the end-Cretaceous
mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today,
climate extremes, in concert with shifts in mean temperature, are expected to
cause the greatest ecological and societal damage \citep{ipcc2012}. But, while
extremes in the physical environment such as wave height, storm severity, and
temperature are frequent events \citep{gaines1993,katz2005}, it remains
unclear the extent to which ecological systems buffer or suffer from black
swans \citep{nunez2012}.

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence across taxa has been elusive. A survey of ecologists
indicated that surprising outcomes of field experiments are far more common
than we assume \citep{doak2008}, and events such as the global invasion of
Argentine ants and the mutation of viruses to infect new hosts could be
considered black swans \citep{nunez2012}. In fact, anecdotes of population
catastrophes are numerous and catastrophes may be the most important element
affecting population persistence \citep{mangel1994}. For marine mammal
populations, we have compelling evidence of catastrophes \citep{gerber2001,
  ward2007}, and recently, heavy-tailed distributions have been discovered in
time-series of marine microbe abundance \citep{segura2013} and time to
extinction for experimental waterflea populations \citep{drake2014}. Despite
these anecotes, as far as we can tell there have been two systematic surveys
for ecological black swans: one with North American breeding birds and the
other with the Global Population Dynamics Database, which uncovered little
clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless
there are methodological challenges to their detection
\citep{allen2001,ward2007}.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
The growth parameter $\lambda$ represents the expected growth rate if $N_t =
1$. The model is density independent if $b = 1$, maximally density dependent if
$b = 0$, and inversely density dependent if $b < 0$. Usually, the process noise
$\epsilon_t$ is modelled as normally distributed, but in our paper we assume it
is drawn from a t distribution with scale parameter $\sigma$ and degrees of
freedom $\nu$. In previous analyses of the GPDD, the Gompertz was most often
identified as the most parsimonious population model that can be fit to these
data \citep{brook2006}.

@@

We investigated possible covariates of heavy-tailed population dynamics
visually and through multilevel modelling. We plotted characteristics of the
time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
life-history characteristics (body length and maximum lifespan obtained from
\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu <
10)$. We formally investigated these relationships by fitting beta
regression multilevel models \citep{ferrari2004}. Beta regression is useful
for continuous response data that range between zero and one (e.g.\
probabilities); we fit our models with a logit link as is common for logistic
regression \citep{ferrari2004}. We incorporated standard deviations around the
means for covariates that were derived from Gompertz model parameter
estimates. To account for broad patterns of phylogenetic relatedness, we
allowed for hierarchical intercepts at the taxonomic class, order, and species
level (Supporting Information). We fit our model in Stan with weakly
informative priors on the coefficients \citep{gelman2008d} and variance
parameters (\citeauthor{gelman2006c} \citeyear{gelman2006c}
\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information).

@@

We found strong, but rare, evidence for black-swan population dynamics. By
defining black-swan dynamics as a greater than $0.5$ probability that $\nu <
10$, our main Gompertz model found evidence for heavy tails most frequently
for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
orders. Accounting for time series length and partially pooling inference
across taxonomic class and order with a multilevel model, there was stronger
evidence for black swans in insect populations than is visually apparent in
Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median
probability of heavy tails were insect orders---however, there was
considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).

@@

@@

Our results clarify previous related analyses. An analysis with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the log-normal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black-swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black
swans in a large proportion of populations. By quantifying the probability of
heavy-tails and allowing for non-stationary time series and density
dependence, our analysis allows for a more nuanced description of the evidence
of ecological black swans. In an earlier study, \citet{keitt1998} described
heavy (power law) tails in breeding bird population abundance. However, this
finding was challenged by \citet{allen2001}, who showed that mixing data
across species could falsely generate heavy tails.
%This rebuttal supports our
%use of the t distribution to represent heavy tails, since a t distribution can
%be represented by a mixture of normal distributions \citep[with the same mean
%and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black-swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions \citep[with the same mean and inverse-gamma-distributed
variances,][]{gelman2014}, we could observe heavy tails if we miss some
underlying mixture of intrinsic processes \citep{allen2001}. That process
might be an aggregation of populations across space, or population diversity,
or an intrinsic change in population variability through time. Extrinsic
forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These
forces could be extreme themselves. For example, extreme climate, predation
from (or competition with) other species experiencing black swans, or sharp
changes in human pressure such hunting, fishing, or habitat destruction might
cause black swans. Alternatively, the interaction of multiple ``normal''
extrinsic forces could give rise to black-swan ecological dynamics. This could
occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if
non-synergistic interactions experience a rare alignment \citep{denny2009}.

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of
policy directions. Ecological resource management can draw from other
disciplines that focus on heavy tails. For example, earthquake preparedness
and response is focussed on black-swan events. Similarly to ecological black
swans, we can rarely predict the specific location and timing of large
earthquakes. But, earthquake preparedness involves spatial planning based on
forecast probabilities to focus early detection efforts and develop disaster
response plans. A similar focus might benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. The presence of black swans also suggests that we develop
management policy that is robust to heavy tails. For instance, setting target
population abundances that are appropriately set back from critical limits may
buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic,
phenotypic, and behavioural diversity may allow some components of populations
to persist when others are affected by disease or extreme environmental forces
\citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and
unexpected, surprising, or counterintuitive ecological dynamics offer a
tremendous opportunity to learn about ecological systems, evaluate when our
models break down, and adjust future management policy \citep{doak2008,
  pine-iii2009, lindenmayer2010}.

Our results suggest a number of research questions related to ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Can we isolate the components of
ecological dynamics that experience black swans by moving from
phenomenological models such as the Gompertz to mechanistic models that, for
example, take into account recruitment dynamics? We expect that greater
insight into the mechanisms and covariates of ecological black swans may be
best obtained through specific geographic and taxonomic subsets of data where
longer time series with low levels of observation error are available
\citep[e.g.][]{segura2013}.

Most importantly, what is the impact of allowing for black swans in forecasts
of ecological risk? In recent decades, ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on ecological
extremes represents the next frontier, particularly in the face of increased
climate extremes \citep{meehl2004,thompson2013,ipcc2012}. Financial analysts
are concerned with the shape of the tails of financial returns because these
directly impact estimates of risk---the probability of a specific magnitude of
undesired event occurring \citep{rachev2008}. A comparable focus in ecology
would increase our estimates of extinction risk, since these would be
disproportionately impacted by downward black-swan events.

@@

We thank J.D. Yeakel and other members of the Earth to Ocean research
group for helpful discussions and comments. We are grateful to the
contributors and maintainers of the Global Population Dynamics Database and to
Compute Canada's WestGrid high-performance computing resources. Silhouette
images were obtained from \texttt{phylopic.org} under Creative Commons
licenses; sources are listed in the Supporting Information. Funding was
provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural
Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada
Research Chairs Program (NKD).

@@

\bibliographystyle{ecologyletters}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
%\bibliography{ms}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}

\caption{Estimates of population dynamics heavy-tailedness for \nuCoefPopN\
  populations of birds, mammals, insects, and fishes. Small values of $\nu$
  ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of
  $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line
  segments represent posterior medians and 50\% / 90\% credible intervals for
  individual populations. Inset plots show probability that $\nu < 10$
  (probability of heavy tails) for populations arranged by taxonomic order and
  sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or
  fewer populations in panel a are omitted for space. Red to yellow points
  highlight populations with a high to moderately high probability of
  heavy-tailed black-swan dynamics.}

\label{fig:nu-coefs}
\end{center}
\end{figure}
\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}

\caption{Potential covariates of heavy-tailed population dynamics (indicated
  by a high probability that $\nu < 10$). Shown are (a--c) parameters from the
  Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b)
  number of time steps, (c) body length, and (d) lifespan. For the Gompertz
  parameters, $\sigma$ refers to the scale parameter of the Student-t
  process-noise distribution, $\lambda$ refers to the expected log abundance
  at the next time step at an abundance of one, $b$ refers to the density
  dependence parameter ($1$ is maximally density independent, $0$ is maximally
  density dependent, and $<0$ is inversely density dependent). Circles
  representing a few sharks, crustaceans, and gastropods are filled in white.
  Median and 90\% credible interval posterior predictions of a beta regression
  multilevel model are shown in panels a and d where there was a high
  probability the slope coefficient was different from zero
  (Fig.~\ref{fig:posteriors}b).}

\label{fig:correlates}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{../analysis/order-posteriors-covariates.pdf}

\caption{Posterior probability distributions from beta regression multilevel
  models. (a) Taxonomic-order-level posterior densities of Pr($\nu < 10$)
  (approximately the probability of heavy tails) after accounting for
  time-series length. Estimates are at the geometric mean of time series
  length across all the data (approximately 27 time steps). Colour shading
  refers to taxonomic class (yellow: fishes, green: insects, purple: birds,
  and red: mammals). Dotted vertical line in panel a indicates the median
  expected Pr($\nu < 10$) from the prior distribution. (b) Main effect
  posterior densities for potential covariates of Pr($\nu < 10$). The beta
  regression models were fit on a logit scale with hierarchical intercepts for
  taxonomic class, taxonomic order, and species. All covariates were
  standardized by subtracting their mean and dividing by twice their standard
  deviation. In both panels, short vertical line segments within the density
  polygons indicate median posterior estimates.}

\label{fig:posteriors}
\end{center}
\end{figure}
